Complex and informative dependencies propagate throughout ecology and can be revealed using copulas

### 2019-05-13 Paper.Rmd
more editing

For the model we examined, causal hypothesis D (Fig. \ref{MasterFigure}) was 
correct that asymmetric tail dependence in evolutionary changes 
can produce similarly asymmetric tail dependence in character distributions across
phylogeny tips (Appendix \ref{SM-evol_mech} for details). This topic is revisited in the Discussion.

@@

Consistent with consequential hypothesis Y, 
left-tail dependent environmental fluctuations increased metapopulation
extinction risk for the spatially 
extended Lewontin-Cohen model (Fig. \ref{fig_extrisk_lcohen_local}). 
This was true for both $N=5$ and $N=25$ habitat patches, and for both local 
(Fig. \ref{fig_extrisk_lcohen_local}) and global (Fig. \ref{SM-fig_extrisk_lcohen_global})
dispersal. We also substantiated using models (Appendix \ref{SM-Consequence_TL_simulation}, \ref{SM-sect:copsurrognd})
a consequential hypothesis Z (Fig. \ref{MasterFigure}), that 
Taylor's law can be strongly influenced by
tail dependence between populations.

@@

<!--what we did and why is it important to study?-->
We showed that non-N copula structure and asymmetric tail dependence are 
common in datasets across multiple sub-disciplines
and levels of organization in ecology (Fig. \ref{MasterFigure}, O), although no published 
research we are aware of has 
taken this structure into account. We hypothesized four 
mechanisms (Fig. \ref{MasterFigure}, A-D) that may cause non-N copula structures. One
mechanism (Fig. \ref{MasterFigure}, A) was via nonlinear environmental influences
on ecological variables. A second mechanism (Fig. \ref{MasterFigure}, B) produces tail dependence via
a Moran effect, whereby dependencies between environmental variables 
beget similar dependencies between ecological variables in different locations. 
We discuss below how commonly these and other
potential mechanisms may opperate. 
We also hypothesized and demonstrated important consequences of 
non-N and tail-dependent copula structure in ecology 
(Fig. \ref{MasterFigure}, X-Z). First (Fig. \ref{MasterFigure}, X), the skewness
of a spatial-average time series is influenced by tail dependencies between its 
constituent time series. Thus right-tail-dependent local time series 
lead to "spiky" spatially averaged time series, with large outbreaks.
These observations could have implications for numerous exploited and pest species
for which the spatial average time series have applied importance.
Second (Fig. \ref{MasterFigure}, Y), metapopulation extinction risk can be accentuated or reduced by tail dependence
properties of environmental variables. We also showed effects of copula 
structure on Taylor's law (Fig. \ref{MasterFigure}, Z), which has been applied in many fields.
In our view, our results make our "overall hypothesis," framed in the Introduction,
reasonable and worthy of broader testing by the research community. We have illustrated 
extensively how such tests might be carried out by performing them ourselves on 
example datasets and models. Further suggestions for future possible tests are below. 
<!--***DAN: make sure you have enough of this, Dan-->
It already seems likely that comprehensive understanding 
of many ecological phenomena is impossible without 
a complete copula characterization of dependence between variables.

Copula approaches are
well developed, statistically [@joe2014_dependence; @nelsen2006_copula], and 
computer implementations exist (e.g., the `copula` and 
`VineCopula` packages in R). Ecologists can start applying these tools
immediately. We created a variety of interrelated randomization procedures
(Appendices \ref{SM-surrog_test}, \ref{SM-Surrog_multi}, \ref{SM-sect:copsurrognd}) 
that build upon existing methods and may be especially useful for ecological applications.

<!--a para on how likely one is to get copula structure via the Liebig mechanism-->
Our first causal mechanism (Fig. \ref{MasterFigure}, A) seems likely to operate commonly.
Liebig's law and the idea of a limiting 
nutrient of growth are dominant paradigms in ecology.
Although we did not systematically test the first two parts of causal 
hypothesis A, many studies document nonlinear influences of environmental
variables on ecological quantities at all levels of organization. 
These parts of A could be tested
straightforwardly for numerous datasets by examing copulas describing dependencies
between environmental and ecological
variables (see below discussion about the green spruce aphid, and 
Appendix \ref{SM-sect:greenspruce}, which illstrate how to do this).
Causal mechanism C, which we also did not test, could be tested straightforwardly
by analyzing copulas of abundances of competing species, across space or
through time.

>>>>

For the model we examined, causal hypothesis D (Fig. \ref{MasterFigure}) was 
correct that asymmetric tail dependence in evolutionary changes 
can produce similarly asymmetric tail dependence in character distributions across
phylogeny tips (Appendix \ref{SM-evol_mech}, Fig. \ref{SM-fig_cop_evo} for details). This topic 
is revisited in the Discussion.

@@

Consistent with consequential hypothesis Y, 
left-tail dependent environmental fluctuations increased metapopulation
extinction risk for the spatially 
extended Lewontin-Cohen model (Fig. \ref{fig_extrisk_lcohen_local}). 
This was true for both $N=5$ and $N=25$ habitat patches, and for both local 
(Fig. \ref{fig_extrisk_lcohen_local}) and global (Fig. \ref{SM-fig_extrisk_lcohen_global})
dispersal. We also substantiated using models (Appendix \ref{SM-Consequence_TL_simulation}, \ref{SM-sect:copsurrognd}, Fig. \ref{SM-fig_spTL})
a consequential hypothesis Z (Fig. \ref{MasterFigure}), that 
Taylor's law can be strongly influenced by
tail dependence between populations.

@@

<!--what we did and why is it important to study?-->
We showed that non-N copula structure and asymmetric tail dependence are 
common in datasets across multiple sub-disciplines
and levels of organization in ecology, although no published 
research we are aware of has 
taken this structure into account. We hypothesized four 
mechanisms that may cause non-N copula structures.
We discuss below how commonly some of our
potential mechanisms may opperate. 
We also hypothesized and demonstrated important consequences of 
non-N copula structure and tail-dependence in ecology. 
For instance, the skewness
of a spatial-average time series is influenced by tail dependencies between its 
constituent time series: right-tail-dependent local time series 
lead to "spiky" spatially averaged time series, with large outbreaks.
Thus tail dependence could have implications for numerous exploited and pest species
for which the spatial average has applied importance.
Extinction risk can be accentuated or reduced by tail dependence
properties of environmental variables. 
In our view, our results make our "overall hypothesis" from the Introduction
reasonable and worthy of broader testing. We illustrated 
how such tests might be carried out by performing them ourselves on 
example datasets and models. Further suggestions for future tests are below.
It already seems likely that comprehensive understanding 
of many ecological phenomena is impossible without 
a complete copula characterization of dependence.

Copula approaches are
well developed [@joe2014_dependence; @nelsen2006_copula], and 
computer implementations exist (the `copula` and 
`VineCopula` packages). Ecologists can start applying these tools
immediately. We created several interrelated randomization procedures
(Appendices \ref{SM-surrog_test}, \ref{SM-Surrog_multi}, \ref{SM-sect:copsurrognd}) 
that build upon existing methods and may be especially useful for ecology.

<!--a para on how likely one is to get copula structure via the Liebig mechanism-->
Our first causal mechanism (Fig. \ref{MasterFigure}, A) seems likely to operate commonly.
Liebig's law and the idea of a limiting 
nutrient of growth are dominant paradigms in ecology.
Although we did not systematically test the first two parts of causal 
hypothesis A, many studies document nonlinear influences of environmental
variables on ecological quantities at all levels of organization. 
These parts of A could be tested
straightforwardly for numerous datasets by examing copulas describing dependencies
between environmental and ecological
variables (see below discussion about the green spruce aphid, and 
Appendix \ref{SM-sect:greenspruce}, which illustrate how to do this).
Causal mechanism C, which we also did not test, could be tested straightforwardly
by analyzing copulas of abundances of competing species, across space or
through time.
